On Thue Numbers

### 2016-07-23 draft.tex
new stuff

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{wrapfig}
\usepackage{caption, subcaption}
\usepackage[style=ieee, backend=bibtex, natbib=true]{biblatex}
\usepackage{standalone}
\author{Theresa Eimer}
\title{On Thue Numbers}
\date{}
\bibliography{library.bib}
\begin{document}
\renewcommand{\figurename}{Fig.}
\maketitle
\thispagestyle{empty}
\newpage
\tableofcontents
\thispagestyle{empty}
\newpage

\section{Introduction}
In 1906, Axel Thue showed the construction of an infinite word where no two adjacent parts are be the same using only three letters \citep{Thue1906}. This concept of nonrepetitiveness has since been applied to other fields, probably most successfully in graph theory. The nonrepetitiveness of graph colorings and the number of colors needed to construct nonrepetitive colorings have been a topic since Alon et. al. introduced the idea in 2002 \citep{Alon2002}. This thesis aims to give an overview over some results of the results, both in vertex- and edge-colorings. 
\newline
First, some general preliminaries in graph and complexity theory will be covered. 
\newline
Then, in the third section, the formal definition of nonrepetitiveness for graph colorings and the complexity of the corresponding decision problem will be shown. 
\newline
The fourth section is about edge-colorings, more specifically Manin's proof that the Thue Number problem is $\Pi^p_2$-complete \citep{Manin2008}. This proof will be examined in detail as it's probably the most important result for nonrepetitive edge-colorings as well as generally applicable, which isn't the case for most results specific to vertex-colorings.
\newline
Section five, therefore, is intended to be a bit of an overview over some relevant results when it comes to nonrepetitive vertex-colorings, though it certainly doesn't aim to be complete. The first part introduces two randomized algorithms, one for finding repetitive paths of a specific length and the other constructing a nonrepetitive coloring. After that, bounds for two classes of graphs that have received particular attention will be presented: planar graphs and trees. 
\newline
Finally, section six will mention other applications and variants of nonrepetitiveness and the Thue Number.
\newpage
\section{Preliminaries}
\subsection{Graph Theory}

A \textit{graph} G is defined as two, for the purpose of this thesis, finite sets V and E, representing the vertices and edges of G respectively, so that E is a subset of $ V ^ 2$, meaning that each edge from E connects two vertices from V \citep[p. 1]{Bollobas1998}. This is denoted as G = (V, E). Hereafter, any graph G will refer to an undirected and unweighted graph unless mentioned otherwise. Thus, edges can be traversed in both directions. Furthermore, all graphs will be simple graphs, not containing multi-edges (more than one edge from a vertex $v_i$ to another vertex $v_j$) or loops (edges starting from and ending at the same vertex) \citep[pp. 3-4]{Gross2013}. 
\newline
\newline
A graph G' = (V', E') is called a \textit{subgraph} of G if it's part of G, that is, if V' is a subset of V and E' is a subset of V \citep[p. 2]{Bollobas1998}.
\newline
\newline
Two vertices $v_i$ and $v_j$ are \textit{adjacent}, if there is an edge $e_k$ that connects the two. Two edges are adjacent if they share one vertex \citep[p. 1-2]{Bollobas1998}. 
\newline
\newline
A \textit{path} (sometimes called walk) in a graph is a sequence of the form 
\begin{center}
	$ W = v_0 e_0 v_1 e_1...e_{n-1} v_n $
\end{center}
where, starting from an initial vertex $v_0$, vertices and edges alternate until the final vertex $v_n$ is reached. In a simple graph, a path can also be represented by either just its vertices \citep[p. 10]{Gross2013} or edges \citep[p. 4]{Bollobas1998}. The length l of a path is the number of edges it contains \citep[p. 5]{Bollobas1998}.
\newline
\newline
For the problems in this thesis, a certain kind of path, the \textit{open path}, is especially important. Open means that no vertex is visited more than once, thus cycles on the path are not possible (see 'trail' in \citep[p. 10]{Gross2013}). An open path through every vertex is called a \textit{Hamiltonian path} \citep[p. 14]{Bollobas1998}.
\newline
\newline
The vertices and edges of Graphs can be colored. An edge-coloring of a graph G is a function that matches each edge to a color contained in C, the set of colors.  A vertex-coloring is a function matching the vertices instead of the edges to the colors in C \citep[pp. 7-8]{Gross2013}. 
\newpage
\begin{wrapfigure}{r}{4cm}
	\includestandalone{pics/clique}
	\caption{8-Clique}
\end{wrapfigure}
\par There are some special graph structures that will become important later on: the \textit{Clique} and the \textit{Hypercube}. A Clique is a subgraph of a graph G, in which every two vertices $v_i$, $v_j$ are adjacent (see figure 1).  A Clique containing k vertices is denoted as $k-Clique$ \citep[pp. 21, 111]{Gross2013}.
\newline
\newline
A Hypercube is an n-dimensional Graph (denoted n-Hypercube) that can be constructed by labeling a set of vertices of size $2^n$ with a different element of ${0, 1}^n$ each and connecting the vertices that differ in one position (see figure 2) \citep[p. 60]{Bollobas1998}. If $n > 1$, the construction of the (n+1)-Hypercube consist of essentially to n-Hypercubes being connected (see figure 3). Both for the Hypercube and Clique, the number of edges connected to a vertex is the same for all vertices, e.g. 7 for a 8-Clique or 4 for a 4-Hypercube.
\newline

@@

To discuss the complexity of the following problems, some background in complexity theory, specifically the \textit{Polynomial Hierarchy}, will be needed. First, however, \textit{P} and \textit{NP}, two specific complexity classes within the hierarchy will be introduced.
\newline
P and NP are similarly defined using Turing machines. P contains all problems that can be solved in polynomial time, that means all problems a Deterministic Turing Machine can decide in p(x) computational steps, p(n) being any polynomial and x the length of the input. The problems in NP can be solved in polynomial time by a Non-deterministic Turing Machine, the definition is analogous \citep{Davis1983}. An alternative way to define NP is that NP contains all problems for which the a solution can be verified in polynomial time. \citep[p. 59]{Goldreich} Its counterpart, the complexity class of coNP, contains all problems where an instance that is not a solution can be verified as such in polynomial time \citep[p. 154-155]{Goldreich}. 
\newline
The levels of the Polynomial Hierarchy contain three complexity classes each: $\Sigma^p_i, \Delta^p_i$ and $\Pi^p_i$ \citep{Meyer1972}. At level 0, that is i = 0, all three of them are equal and P: 
\begin{center}
	$\Sigma^p_0 = \Delta^p_0 = \Pi^p_0 = P $ \citep{Stockmeyer1976}
\end{center}  
The further levels are recursively defined using \textit{oracles}. If a Turing Machine can use an oracle, it means that in a single computational step, a L oracle will decide if a string on the Turing Machine's tape is contained in L. With that in mind, the definitions go as follows:
\newline
\newline
	$\Sigma^p_{i+1} = \{$ L | L is accepted by a Non-deterministic Turing Machine with a L' oracle for some L' $\epsilon \Sigma^p_i\}$
\newline
\newline
	$\Pi^p_{i+1} = \{$ L | the complement of L is accepted by a Non-deterministic Turing Machine with a L' oracle for some L' $\epsilon \Sigma^p_i\}$
\newline
\newline
	$\Delta^p_{i+1} = \{$ L | L is accepted by a Deterministic Turing Machine with a L' oracle for some L' $\epsilon \Sigma^p_i\}$
\newline 
\newline
P and NP are of course included here, $\Delta^p_1 = P$ and $\Sigma^p_1 = NP$ \citep{Meyer1972}. $\Pi^p_1 $ is called co-NP.
\newline
$\Sigma^p_i$, $\Pi^p_i$ and $\Delta^p_i$ are related in the following way:
\begin{center}
	$\Sigma^p_i \cup \Pi^p_i \subseteq \Delta^p_{i+1} \subseteq \Sigma^p_{i+1} \cap \Pi^p_{i+1}$
\end{center}
It's not yet decided if these inclusions are proper, but they give an idea of what the Polynomial Hierarchy looks like \citep{Stockmeyer1976}.
\newline
One possible way to decide which complexity class a problem belongs to, is the reduction from a problem with known complexity. Let L and Q be two languages, then Q is reducible to L if there is a function f(n) that runs in polynomial time so that for any x in Q, f(x) is in L. This is written as $Q \leq_p L$. If such a function exists, L is at least as complex as Q \citep{Davis1983}. 
\newline
Three problems that will be reduced from later on in this thesis are SAT, 3SAT and $\forall \exists 3SAT$. The later two are special cases of SAT, the boolean satisfiability problem. SAT is the language containing all boolean formulas in conjunctive normal form that are satisfiable. SAT is in NP and in fact NP-hard, meaning that every problem in NP can be reduced to SAT. Both of those attributes combined make it NP-complete. \citep[p. 338-346]{Davis1983} 3SAT is a subset of SAT with all formulas with 3 or less literals per clause and is also NP-complete. \citep[p. 347]{Davis1983} $\forall \exists 3SAT$ finally is the satisfiability problem for formulas $\varphi$ with 3 or less literals per clause that are of the form $\forall (x_1, ..., x_i) \exists (y_1, ..., y_j) \varphi(x_1, ..., x_i, y_1, ..., y_j)$. It's $\Pi^p_2$-complete \citep{Schaefer2002}.
\section{Nonrepetitive graph colorings}
\subsection{Nonrepetitiveness}
Before talking about Thue number type problems, it makes sense to look at non-repetitiveness first, as it is the defining concept behind the  Thue number. Nonrepetitive here means square-free. This can be applied to different fields like graph theory, geometry or number theory \citep{Grytczuk2008}, but it was first introduced in the process of constructing a square-free word by Axel Thue in 1906. A word w is square-free if there is no part x of the word that occurs two times in a row. Let the words a and b each consist of three different substrings, v, x and y, then a = vxxy is not square-free, but b = vxy is \citep{Thue1906}.

@@

In graph theory, nonrepetitiveness is relevant with respect to nonrepetitive vertex- \citep{Marx2009} and edge-colorings of graphs. A coloring e of a graph G is nonrepetitive, if there is no open path whose vertex or edge colors contain a color sequence x that is repeated two or more times in a row, similar to the definition of nonrepetitiveness for words. To put it differently: a coloring of a graph is nonrepetitive if the coloring of every possible open path on the graph is nonrepetitive \citep{Alon2002}. To give a baseline for the complexity of further problems concerning nonrepetitive graph colorings, first the complexity of determining whether a vertex- or edge coloring is nonrepetitive will be shown.

@@

The problem of deciding whether a vertex-coloring of a graph G(V, E) is nonrepetitive has been proven to be coNP-complete by reducing from the Hamiltonian path problem, which is NP-complete. The idea behind Marx's and Schäfer's proof \citep{Marx2009a} is to construct a graph H from G with two parts, the second one having a path for every possible path in G that is as long a Hamiltonian path would be (n vertices with n = |E|), and the first part representing the vertices of G with a subgraph each. H is colored in such a way that a repetitively colored path would have to go through both parts with the coloring from the first path repeating itself in the second path. If a path through the second half of H, however, corresponds to a path in G that is not a Hamiltonian path but a path that visits at least one vertex more than once, the subgraph for that vertex in the first half of H would have to also be visited more than once - but that would mean that the path through the first half is not an open path. Thus H is colored nonrepetitively only if G doesn't have a Hamiltonian path. This implies that the problem of deciding if a vertex-coloring is nonrepetitive is coNP-hard.

@@

The construction of H goes as follows: for the first part, for every vertex $i \leq n$ (n = |E|) of G, a $K_{2, n}$ graph (fig. 4) is added. It consists of two vertices, colored a and b respectively, with n vertices between them, each connected to both the a- and b-vertex. As one vertex is added for each vertex in G, the vertices are colored $c_{i,j}$, i being the vertex in G for which the subgraph was added and j ranging from 1 to n. The b-colored vertices are now connected to  a  vertex, z, colored c, that marks the transition between the two parts of H. Lastly, to connect the different subgraphs, for every b-vertex, $d_{i,j}$-colored vertices are added. A $d_{i, j}$ vertex is connected to the b-vertex of the subgraph with the same value for i and all a-vertices for j between 1 and n where $j \neq i$.  So now all a and b vertices are connected to all other $K_{2,n}$ subgraphs of the first part of H.

@@

	\caption{\\Graph F}

@@

	\caption{The first part of H constructed for F; \\The marked edges are connected to a $d_{i,j}$-colored vertex}

@@

 For the second part, for every combination of i and j ($1 \leq i, j \leq n$), a path containing three vertices colored a, $c_{i,j}$ and b respectively are added to H. Where j = 1, the a-colored vertex is connected to the middle vertex z. Between the b-colored vertices for j = 1 and the a-colored vertices for j = 2, $d_{i,j}$-colored vertices are added. i here is the same as in three vertex path to which the new vertex becomes connected via the b-vertex, j the number of every vertex the i-th vertex in G is connected to by an edge in E. Those d-colored vertices are then connected to the a-vertex of the three vertices path corresponding to the j value in their color. This way, the connections in G are mimicked in the structure of H. This is repeated for every j until all the three vertex subgraphs have been connected. The last step is to connect the b-vertices of the paths where j = n to a c-colored end vertex z'. 

@@

	\caption{The whole of H constructed for F \\ Edges connected to vertices colored $d_{i,j}$ are marked}

@@

Now H has a first part that allows any path through the graph to be traced through the coloring, a vertex of a different color separating the two halves, and a second part whose connections are dependent on the edges of G. In the first and second path, a and b are the only non-unique colors and vertices colored a or b are never adjacent, so a repetitive path has run through both parts, that means it needs to pass the c-colored vertex z, and because of this also needs to reach the end vertex as it is the only other vertex colored with c. Thus, the only possible square path would look like this: w = QzQ'z' with Q and Q' and of course z and z' having the same coloring. To get from vertex z to vertex z', w has to pass through n vertices colored $c_{i,j}$ for some i,j, visiting the same colors in the first part of H. As all vertices colored $c_{i,j}$ with the same value for i are contained in the same subgraph of the first half of H, w can only visit one $c_{i,j}$ colored vertex for each value of i. This is only possible if G has a Hamiltonian path, because then, by construction of the second part, there is a path through the second half of H that doesn't visit any two vertices with the same value for i. The coloring of this path can simply be copied for a path through the first part. If G doesn't have a Hamiltonian path, there is no path 
through the second part of H without having to visit two different vertices with 
the same value for i, but to get the same coloring for the first part, the path \begin{wrapfigure}{l}{4cm}

@@

there would have to pass through one of the $K_{2,n}$ subgraphs twice, therefore it is not an open path. F is an example of a graph without a Hamiltonian path, so it will not be possible to find a path of the form w = QzQ'z', the example below from Marx's and Schäfer's paper, however, does contain a square path. The graph it's based on is F'. 

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage[normalem]{ulem} 
\usepackage{wrapfig}
\usepackage{caption, subcaption}
\usepackage{todonotes}
\usepackage[style=ieee, backend=bibtex, natbib=true]{biblatex}
\usepackage{standalone}
\author{Theresa Eimer}
\title{On Thue Numbers}
\date{}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{claim}{Claim}

\bibliography{library.bib}
\begin{document}
\renewcommand{\figurename}{Fig.}
\maketitle
\thispagestyle{empty}
\newpage
\tableofcontents
\thispagestyle{empty}
\newpage

\section{Introduction}
In 1906, Axel Thue showed the construction of an infinite word where no two adjacent parts are be the same using only three letters \citep{Thue1906}. This concept of nonrepetitiveness has since been applied to other fields, probably most successfully in graph theory. The nonrepetitiveness of graph colorings and the number of colors needed to construct nonrepetitive colorings have been a topic since Alon et. al. introduced the idea in 2002 \citep{Alon2002}. This thesis aims to give an overview over some results of the results, both in vertex- and edge-colorings. 
\newline
First, some general preliminaries in graph and complexity theory will be covered. 
\newline
Then, in the third section, the formal definition of nonrepetitiveness for graph colorings and the complexity of the corresponding decision problem will be shown. 
\newline
The fourth section is about edge-colorings, more specifically Manin's proof that the Thue Number problem is $\Sigma^p_2$-complete \citep{Manin2008}. This proof will be examined in detail as it's probably the most important result for nonrepetitive edge-colorings as well as generally applicable, which isn't the case for most results specific to vertex-colorings.
\newline
Section five, therefore, is intended to be a bit of an overview over some relevant results when it comes to nonrepetitive vertex-colorings, though it certainly doesn't aim to be complete. The first part introduces two randomized algorithms, one for finding repetitive paths of a specific length and the other constructing a nonrepetitive coloring. After that, bounds for two classes of graphs that have received particular attention will be presented: planar graphs and trees. 
\newline
Finally, section six will mention other applications and variants of nonrepetitiveness and the Thue Number.
\newpage
\section{Preliminaries}
\subsection{Graph Theory}

\begin{definition}
A graph $G$ is defined as two finite sets V and E, representing the vertices and edges of $G$ respectively, so that $E$ is a subset of $ V ^ 2$, meaning that each edge from $E$ connects two vertices from $V$ \citep[p. 1]{Bollobas1998}.
\end{definition} 
This is denoted as $G = (V, E)$. Hereafter, any graph $G$ will refer to an undirected and unweighted graph unless mentioned otherwise. Thus, edges can be traversed in both directions. Furthermore, all graphs will be simple graphs, not containing multi-edges (more than one edge from a vertex $v_i$ to another vertex $v_j$) or loops (edges starting from and ending at the same vertex) \citep[pp. 3-4]{Gross2013}. 
\begin{definition}
A graph $G' = (V', E')$ is called a subgraph of $G$ if $V'$ is a subset of $V$ and $E'$ is a subset of $E$ \citep[p. 2]{Bollobas1998}.
\end{definition}
\begin{definition}
Two vertices $v_i$ and $v_j$ are \textit{adjacent}, if there is an edge $e_k$ that connects the two. Two edges are adjacent if they share one vertex \citep[p. 1-2]{Bollobas1998}. 
\end{definition}
\begin{definition}
A path on a graph is a sequence of the form 
\begin{center}
	$ W = v_0 e_0 v_1 e_1...e_{n-1} v_n $
\end{center}
where, starting from an initial vertex $v_0$, vertices and edges alternate until the final vertex $v_n$ is reached.
\end{definition}
In a simple graph, a path can also be represented by either just its vertices \citep[p. 10]{Gross2013} or edges \citep[p. 4]{Bollobas1998}. The length $l$ of a path is the number of edges it contains \citep[p. 5]{Bollobas1998}.
\begin{definition}
An open path is a path that visits no vertex more than once.
\end{definition} Thus cycles on the path are not possible (see 'trail' in \citep[p. 10]{Gross2013}). Open paths are sometimes called walks. 
\begin{definition}An open path through every vertex is called a Hamiltonian path \citep[p. 14]{Bollobas1998}.
\end{definition}
\begin{definition}
An edge-coloring of a graph $G$ is a function that matches each edge to a color contained in $C$, the set of colors.  A vertex-coloring is a function matching the vertices instead of the edges to the colors in $C$ \citep[pp. 7-8]{Gross2013}.
\end{definition} 
\newpage
\begin{wrapfigure}{r}{4cm}
	\includestandalone{pics/clique}
	\caption{8-Clique}
\end{wrapfigure}
\par There are some special graph structures that will become important later on: the \textit{Clique} and the \textit{Hypercube}. 
\begin{definition}
A Clique is a subgraph of a graph G, in which every two vertices $v_i$, $v_j$ are adjacent (see figure 1).  A Clique containing k vertices is denoted as $k-Clique$ \citep[pp. 21, 111]{Gross2013}.
\end{definition}
\begin{definition}
A Hypercube is an $n$-dimensional Graph (denoted n-Hypercube) that can be constructed by labeling a set of vertices of size $2^n$ with a different element of $\{ 0, 1 \}^n$ each and connecting the vertices that differ in one position (see figure 2) \citep[p. 60]{Bollobas1998}.
\end{definition}
If $n > 1$, the construction of the $(n+1)$-Hypercube consist essentially of two $n$-Hypercubes being connected (see figure 3). Both for the Hypercube and Clique, the number of edges connected to a vertex is the same for all vertices, e.g. 7 for a 8-Clique or 4 for a 4-Hypercube.
\newline

@@

To discuss the complexity of the following problems, some background in complexity theory, specifically the \textit{Polynomial Hierarchy}, will be needed. First, however, $P$ and $NP$, two specific complexity classes within the hierarchy will be introduced.

@@

$P$ and $NP$ are similarly defined using Turing machines. 
\begin{definition}
$P$ contains all problems that can be solved in polynomial time, that means all problems a Deterministic Turing Machine can decide in $p(x)$ computational steps, $p(n)$ being any polynomial and $x$ the length of the input.
\end{definition} 
\begin{definition}
$NP$ contains all problems that can be solved in polynomial time by a Non-deterministic Turing Machine \citep{Davis1983}. 
\end{definition} 
An alternative way to define $NP$ is that $NP$ contains all problems for which the a solution can be verified in polynomial time. \citep[p. 59]{Goldreich} Its counterpart, the complexity class of $coNP$, contains all problems where an instance that is not a solution can be verified as such in polynomial time \citep[p. 154-155]{Goldreich}. 
The levels of the Polynomial Hierarchy contain three complexity classes each: $\Sigma^p_i, \Delta^p_i$ and $\Pi^p_i$ \citep{Meyer1972}. At level 0, that is $i = 0$, all three of them are equal and P: 
\begin{definition}
	$\Sigma^p_0 = \Delta^p_0 = \Pi^p_0 = P $ \citep{Stockmeyer1976}
\end{definition}  
The further levels are recursively defined using \textit{oracles}. If a Turing Machine can use an oracle, it means that in a single computational step, a $L$ oracle will decide if a string on the Turing Machine's tape is contained in $L$. With that in mind, the definitions go as follows:
\begin{definition}
	$\Sigma^p_{i+1} = \{$ $L$ | $L$ is accepted by a Non-deterministic Turing Machine with a $L'$ oracle for some $L'$ $\epsilon \Sigma^p_i\}$
\end{definition}
\begin{definition}
	$\Pi^p_{i+1} = \{$ $L$ | the complement of $L$ is accepted by a Non-deterministic Turing Machine with a $L'$ oracle for some $L'$ $\epsilon \Sigma^p_i\}$
\end{definition}
\begin{definition}
	$\Delta^p_{i+1} = \{$ $L$ | $L$ is accepted by a Deterministic Turing Machine with a $L'$ oracle for some $L'$ $\epsilon \Sigma^p_i\}$
\end{definition}
$P$ and $NP$ are of course included here, $\Delta^p_1 = P$ and $\Sigma^p_1 = NP$ \citep{Meyer1972}. $\Pi^p_1 $ is called $coNP$.
$\Sigma^p_i$, $\Pi^p_i$ and $\Delta^p_i$ are related in the following way:
\begin{center}
	$\Sigma^p_i \cup \Pi^p_i \subseteq \Delta^p_{i+1} \subseteq \Sigma^p_{i+1} \cap \Pi^p_{i+1}$
\end{center}
It's not yet decided if these inclusions are proper, but they give an idea of what the Polynomial Hierarchy looks like \citep{Stockmeyer1976}.
One possible way to decide which complexity class a problem belongs to, is the reduction from a problem with known complexity. 
\begin{definition} Let $L$ and $Q$ be two languages, then $Q$ is reducible to $L$ if there is a function $f(n)$ that runs in polynomial time so that for any $x$ in $Q$, $f(x)$ is in $L$. 
\end{definition} 
This is denoted as $Q \leq_p L$. If such a function exists, $L$ is at least as complex as Q \citep{Davis1983}. 
Three problems that will be reduced from later on in this thesis are $SAT$, $3SAT$ and $\forall \exists 3SAT$. The later two are special cases of $SAT$, the boolean satisfiability problem. 
\begin{definition} 
$SAT$ is the language containing all boolean formulas in conjunctive normal form that are satisfiable. 
\end{definition} 
$SAT$ is in $NP$ and in fact $NP-hard$, meaning that every problem in $NP$ can be reduced to $SAT$. Both of those attributes combined make it $NP-complete$. \citep[p. 338-346]{Davis1983} 
\begin{definition} $3SAT$ is a subset of $SAT$ with all formulas with 3 or less literals per clause.
\end{definition} 
$3SAT$ is also $NP-complete$. \citep[p. 347]{Davis1983} 
\begin{definition}
$\forall \exists 3SAT$ finally is the satisfiability problem for formulas $\varphi$ with 3 or less literals per clause that are of the form \\ $\forall (x_1, ..., x_i) \exists (y_1, ..., y_j) \varphi(x_1, ..., x_i, y_1, ..., y_j)$.
\end{definition} 
It's $\Pi^p_2$-complete \citep{Schaefer2002}.

\newpage
\section{Nonrepetitive graph colorings}
\subsection{Nonrepetitiveness}
Before talking about Thue number type problems, it makes sense to look at non-repetitiveness first, as it is the defining concept behind the  Thue number. Nonrepetitive here means square-free. This can be applied to different fields like graph theory, geometry or number theory \citep{Grytczuk2008}, but it was first introduced in the process of constructing a square-free word by Axel Thue in 1906. 
\begin{definition} 
A word $w$ is square-free if there is no part $x$ of the word that occurs two times in a row.
\end{definition} 
Let the words $a$ and $b$ each consist of three different substrings, $v$, $x$ and $y$, then $a = vxxy$ is not square-free, but $b = vxy$ is \citep{Thue1906}.
\newline
In graph theory, nonrepetitiveness is relevant with respect to nonrepetitive vertex- \citep{Marx2009} and edge-colorings of graphs. 
\begin{definition} 
A coloring $e$ of a graph $G$ is nonrepetitive, if there is no open path whose vertex or edge colors contain a color sequence $x$ that is repeated two or more times in a row.
\end{definition} 
To put it differently: a coloring of a graph is nonrepetitive if the coloring of every possible open path on the graph is nonrepetitive \citep{Alon2002}. To give a baseline for the complexity of further problems concerning nonrepetitive graph colorings, first the complexity of determining whether a vertex- or edge coloring is nonrepetitive will be shown.

@@

\begin{theorem}
Deciding whether a vertex-coloring of a graph is nonrepetitive is coNP-complete. \citep{Marx2009a}
\end{theorem}
The problem of deciding whether a vertex-coloring of a graph $G(V, E)$ is nonrepetitive has been proven to be $coNP-hard$ by reducing from the Hamiltonian path problem, which is $NP-complete$. The idea behind Marx's and Schäfer's proof is to construct a graph H from G with two parts, the second one having a path for every possible path in G that is as long a Hamiltonian path would be ($n$ vertices with $n = |E|$), and the first part representing the vertices of $G$ with a subgraph each. $H$ is colored in such a way that a repetitively colored path would have to go through both parts with the coloring from the first path repeating itself in the second path. If a path through the second half of $H$, however, corresponds to a path in $G$ that is not a Hamiltonian path but a path that visits at least one vertex more than once, the subgraph for that vertex in the first half of $H$ would have to also be visited more than once - but that would mean that the path through the first half is not an open path. Thus $H$ is colored nonrepetitively only if $G$ doesn't have a Hamiltonian path. This implies that the problem of deciding if a vertex-coloring is nonrepetitive is $coNP-hard$.

@@

The construction of $H$ goes as follows: for the first part, for every vertex $i \leq n$ ($n = |E|$) of $G$, a $K_{2, n}$ graph (fig. 4) is added. It consists of two vertices, colored $a$ and $b$ respectively, with $n$ vertices between them, each connected to both the $a$- and $b$-vertex. As one vertex is added for each vertex in $G$, the vertices are colored $c_{i,j}$, $i$ being the vertex in $G$ for which the subgraph was added and $j$ ranging from 1 to $n$. The $[b]$ vertices are now connected to $[a]$ vertex, $z$, colored $c$, that marks the transition between the two parts of $H$. Lastly, to connect the different subgraphs, for every $[b]$ vertex, $d_{i,j}$-colored vertices are added. A $d_{i, j}$ vertex is connected to the $[b]$ vertex of the subgraph with the same value for $i$ and all $[a]$ vertices for $j$ between 1 and $n$ where $j \neq i$.  So now all $[a]$ and $[b]$ vertices are connected to all other $K_{2,n}$ subgraphs of the first part of $H$.

@@

	\caption{\\Graph $F$}

@@

	\caption{The first part of $H$ constructed for $F$; \\The marked edges are connected to a $d_{i,j}$-colored vertex}

@@

For the second part, for every combination of $i$ and $j$ ($1 \leq i, j \leq n$), a path containing three vertices colored a, $c_{i,j}$ and $b$ respectively are added to $H$. Where $j = 1$, the $[a]$ vertex is connected to the middle vertex $z$. Between the $[b]$ vertices for $j = 1$ and the $[a]$ vertices for $j = 2$, $d_{i,j}$-colored vertices are added. $i$ here is the same as in three vertex path to which the new vertex becomes connected via the $[b]$ vertex, $j$ the number of every vertex the $i$-th vertex in $G$ is connected to by an edge in $E$. Those $[d]$ vertices are then connected to the $[a]$ of the three vertices path corresponding to the $j$ value in their color. This way, the connections in $G$ are mimicked in the structure of $H$. This is repeated for every $j$ until all the three vertex subgraphs have been connected. The last step is to connect the $[b]$ vertices of the paths where $j = n$ to a $[c]$ end vertex $z'$. 

@@

	\caption{The whole of $H$ constructed for $F$ \\ Edges connected to vertices colored $d_{i,j}$ are marked}

@@

Now $H$ has a first part that allows any path through the graph to be traced through the coloring, a vertex of a different color separating the two halves, and a second part whose connections are dependent on the edges of $G$. In the first and second path, $a$ and $b$ are the only non-unique colors and vertices colored $a$ or $b$ are never adjacent, so a repetitive path has run through both parts, that means it needs to pass the $c$-colored vertex $z$, and because of this also needs to reach the end vertex as it is the only other vertex colored with $c$. Thus, the only possible square path would look like this: $w = QzQ'z'$ with $Q$ and $Q'$ and of course $z$ and $z'$ having the same coloring. To get from vertex $z$ to vertex $z'$, $w$ has to pass through $n$ vertices colored $c_{i,j}$ for some $i,j$, visiting the same colors in the first part of $H$. As all vertices colored $c_{i,j}$ with the same value for $i$ are contained in the same subgraph of the first half of $H$, $w$ can only visit one $c_{i,j}$ colored vertex for each value of $i$. This is only possible if $G$ has a Hamiltonian path, because then, by construction of the second part, there is a path through the second half of $H$ that doesn't visit any two vertices with the same value for $i$. The coloring of this path can simply be copied for a path through the first part. If $G$ doesn't have a Hamiltonian path, there is no path 
through the second part of $H$ without having to visit two different vertices with the same value for $i$, but to get the same coloring for the first part, the path there would have to pass through one of the $K_{2,n}$ subgraphs twice, therefore it is not an open path. $F$ is an example of a graph without a Hamiltonian path, so it will not be possible to find a path of the form $w = QzQ'z'$, the example below from Marx's and Schäfer's paper, however, does contain a square path. The graph it's based on is $F'$. 
\todo{Add pic}

@@

Obviously, the construction of $H$ takes a polynomial number of steps depending on the number of vertices in $G$. So the reduction works, making the problem of deciding the nonrepetitiveness of a vertex-coloring $coNP-hard$. 
\begin{claim}
Deciding whether a vertex-coloring of a graph is nonrepetitive is in $coNP$.
\end{claim} 
Let $G'$ be a graph with an arbitrary vertex-coloring. A Turing machine can non-deterministically pick one path on $G'$ and simply test every possible pattern from every vertex to see if it repeats itself. There are $\frac{n - i}{2}$ possible patterns for a path of length $n$, starting from the $i$-th vertex that could be repeated. Going through them all would take $n \cdot \frac{n - i}{2}$ steps. Thus, a Turing machine can non-deterministically decide if the given coloring is a negative instance of the problem, containing a path that is repetitively colored,  in O($|V^2|$) steps and that is precisely the definition of $coNP$. 
So the problem is $coNP-complete$.

\begin{theorem}
Deciding whether an edge-coloring is nonrepetitive is coNP-complete \citep{Manin2008}.
\end{theorem}
The corresponding problem for edge-colorings is in the same complexity class and the proof is somewhat similar, they are however not directly related and one can't be inferred from the other \citep{Marx2009a}. The reason Manin's proof might seem familiar is that it's also based on the idea of a path with two parts that are colored identically to form the only possible repetitive colored path. The problem he reduces from is $3SAT$, corresponding to which a graph $G$ is constructed. $G$ has a so called snout, a sequence of vertices and edges with the sole purpose of providing the colors for the second part to mimic. The second part of $G$ then consist of a variable gadget for each variable that splits into a positive and negative path variant and a clause gadget for each clause. The clause gadgets are connected to the variable gadgets in the same way the variables occur in the clause; if a variable is negated, the clause gadget is connected to a vertex in the negative path of the variable gadget, if it isn't, the clause gadget is connected to the positive path. As a result, the only way to pass through the second half of G with an open path, is find an assignment satisfying the formula, traversing the variable gadgets opposite to that assignment and then the clause gadgets corresponding to it. If there is no satisfying assignment, at least one of the edges from the clause gadgets will need to pass through a vertex in the variable gadget that has already been visited. Thus a coloring in $G$ is only repetitive if the $3SAT$ problem it's constructed from is satisfiable - and nonrepetitive if it isn't.
To construct $H$, a few numbers from the $3SAT$ problem $f$ are needed. Let $M$ be the maximum of the  number of instances of all variables, meaning the number of instances of the most common variable with $x$ and and $\neg x$ being counted separately, $n$ the number of variables and $m$ the number of clauses in $f$. Then first, $n$ variable gadgets with the following structure are added: each variable gadget begins with a vertex $b_i$ and then splits into a positive and negative path, both of containing $M$ vertices. The paths are both connected to the starting vertex of the next variable gadget, $b_{i+1}$, except for the last gadget which is connected to a vertex $c$. Now the clause gadgets are added, one vertex for each clause in $f$ and one vertex $d$. The first clause gadget is connected to $c$ and to a vertex in every variable gadget corresponding to a variable the clause contains. If the the clause is $(x \vee y \vee \neg z)$, the clause gadget will be connected to one vertex in the positive path of $x$, one vertex in the positive path of $y$ and one vertex in the negative path of $z$. Those vertices in turn are connected to the next clause gadget, unless there are no further clause gadgets, then they're connected to $d$. A vertex in the variable gadgets can't be connected to more than two clause gadgets. This is possible because the paths in the variable gadgets each have $M$ vertices, ensuring that no variable has more instances than distinct vertices in its corresponding path. The snout is fairly simple now, it's merely a sequence of $(M+1)n+2m+1$ \footnote{In \citep{Manin2008} the number is $Mn+2m$, however, the snout is too small this way} vertices, with every vertex $a_i$ connected to $a_{i+1}$ and $a_{(M+1)n+2m+1}$ connected to the first  vertex of the first variable gadget, $b_1$. An example for $f(x, y, z) = (x \vee y) \wedge \neg x \wedge \neg y$ is given in figure 9. 
	\caption{$G$ constructed from $f$ \\edge gadgets are highlighted}
